Impacts of Isolation by Distance on Demographic Modeling and Association Studies

### 2019-03-15 ms/spaceness.tex
adding sampling to ms

\begin{abstract}
Individuals exist in continuous space, but standard models in population genetics are based on discrete randomly-mating populations exchanging migrants. As the availability of population-level, whole-genome data allows inference of increasingly fine-scale patterns of ancestry in many species, models incorporating realistic demographic and spatial processes are needed to accurately describe spatial structure and control for its impacts on analyses of selection and demography. Here we implement a forward-time simulation of evolution in continuous space and use it to study the impacts of dispersal and density-dependent competition on population genetic summary statistics, demographic inference, and association studies. Low dispersal slows the geographic spread of ancestry and increases branch lengths corresponding to mid-frequency alleles because of slow coalescence among distant groups of individuals. As a result demographic inference from the site frequency spectrum is biased towards inferring a turbulent past demographic history and a recent decline in populations when neighborhood sizes are below 20. In association analyses spatially correlated environments cause genome-wide signals of genetic association with purely environmental phenotypes, and this bias is only partially corrected by regressing out principal components positions during the analysis. We find that association analyses in continuous populations are most biased when environmental effects are spatially discrete, but that corrections for clinal phenotypes can also fail under high dispersal because PCA describes less total variation in the data when isolation by distance is weak.

\end{abstract}

@@

In most sexual organisms individuals mate with other nearby individuals, leading to a positive correlation between genetic and geographic distances. This pattern of "isolation by distance" \citep{Wright1943} is one of the most widely replicated empirical findings in population genetics \citep{Chen2017,Jay2012,Sharbel2000}, but is only heuristically approximated by existing forward-time models of molecular evolution based on discrete demes exchanging migrants. As the availability of population-level whole-genome data allows inference of increasingly fine-scale patterns of ancestry in many species, models incorporating biologically realistic demographic and spatial processes are needed to accurately describe continuous spatial structure and control for its impacts on analyses of selection and demography. This situation is particularly clear in human population genetics, where individual ancestries are often interpreted as mixtures of discrete populations in part because of modeling assumptions underlying discrete clustering algorithms and association analyses are often strongly dependent on statistical corrections for population stratification \citep{Berg2018}.

The best-studied approaches to population genetics in continuous space were developed by \citep{Wright1943} and \citep{Malecot1948}, who derived expressions for pairwise genetic differentiation in continuous space assuming Poisson distributed numbers of offspring and independent dispersal among individuals. However, \citep{Felsenstein1975} showed that these assumptions are incompatible -- over time, a population meeting them will clump into a small number of geographic clusters occupying only a part of the available range. (NOTE: where to put Rousset et al. lit?)

One method for modeling populations in continuous space is then to assume the existence of a grid of discrete randomly-mating populations connected by migration, which prevents clustering by forcing all regions to be occupied in every generation. Though good approximations of continuous structure given high dispersal (Barton(?)), these models are not truly continuous and limit investigation of spatial structure below the level of the deme. An alternative approach is to model the geographic spread of ancestry backwards in time through a diffusion approximation -- an approach that has recently made significant progress in modeling both dispersal and demographic parameters \citep{Barton2010,Kelleher2014,Ringbauer2017,Ringbauer2018}. However, these models currently lack a well-described forward-time analog that produces biologically realistic spatial patterns.

A direct approach to the clustering problem of classical forward-time models is to incorporate density dependence by scaling individual fitness to local population density. This shifts reproductive output towards regions of low-density, preventing populations from clustering in space over time. This approach was taken previously by \citep{Doebeli2003} who used an individual based model with continuous space and density dependent fitness to study the probability of speciation along continuous environmental gradients. (**Sentence referencing other continuous IBM's with density dependence**). However to our knowledge all previous implementations of continuous space models have focused on individuals or at most a few genotypes as the unit of analysis, which limits our ability to investigate the impacts of continuous space on genome-wide genetic variation as is now routinely sampled from real organisms.

Here we describe an implementation of an individual-based model in continuous space that incorporates overlapping generations, a Gaussian dispersal kernel, and density-dependent fitness, and scales to chromosome-scale alignments across tens of thousands of individuals. We use this model to describe the impacts of variation in dispersal distance on a variety of summary statistics typically analyzed under discrete population models, and apply a machine-learning regression approach to evaluate a new class of summary statistics based on correlations between geographic distance and moments of the distribution of pairwise haplotype block lengths as estimators of dispersal distance. We then examine how the fine-scale spatial structures occurring under limited dispersal impact demographic inference from the site frequency spectrum; finding that inflation of branch lengths for mid-frequency alleles associated with slow coalescence of distant individuals **TBD**.

Last, we use our model to study the impacts of isolation by distance in continuous space on genome-wide association studies (GWAS). Variation in ancestry proportions between case and control cohorts is expected to cause inflation of test statistics at SNPs with different allele frequencies among populations (CITE). In addition, because most phenotypes are influenced by the environment and environmental factors are often spatially correlated, subtle spatial structure within and among populations can lead to inflated effect size estimates in GWAS of quantitative traits conducted on samples from large geographic regions, and this effect is particularly prominent in low-frequency alleles \citep{Mathieson2012}. For example, imagine a population of flowering plants occurring over a large latitudinal range with populations subject to isolation by distance. Flowering time is strongly dependent on photoperiod and temperature, both of which typically vary with latitude. If we were to conduct a GWAS for flowering time without correcting for population structure we would find that all SNPs with allele frequencies varying clinally with latitude were significantly associated with flowering time, even if flowering time was determined purely by the environment.

A common method of correcting for population structure in GWAS is to run a principal components analysis on the genotype matrix and include PC coordinates as covariates in a linear regression for each SNP \citep{Price2006}. Mixed model GWAS employs a different statistical framework but takes a similar approach by incorporating a kinship matrix describing pairwise genetic relatedness among samples as a random effect in a linear mixed model \citep{Yu2006,Kang2008}. In both approaches the key question is whether or not the ancestry components of the model adequately describe "background" levels of genetic variation in the sample relative to any confounding effects of ancestry and phenotype. Recent studies in European and British populations \citep{Kerminen2018,Berg2018,Sohail2018} have found evidence of residual population stratification in effect size estimates from large-scale GWAS employing both linear and mixed model approaches, suggesting that existing structure corrections may be inadequate. For traits controlled by a few large effect loci this confounding effect is likely to be relatively small, but for analyses of complex traits with thousands of putatively associated SNPs the potential for confounding is high. Here we simulate a range of spatially correlated phenotypes for simulated individuals and seek to identify conditions under which existing GWAS structure corrections fail.  
\section{Materials and Methods}
\label{sec:materials:methods}

@@

To isolate spatial effects from other components of the model such as overlapping generations, increased variance in reproductive success, or density-dependent fitness, we  also implemented a second version of our simulation in which mates are selected at random and offspring disperse to a random location on the landscape. 
For all simulations the full history of the genomes of final-generation individuals were stored as tree sequences \citep{Kelleher2018}.

@@

We ran 400 simulations for the spatial and random-mating SLiM models on a square landscape of size 50 with per-unit carrying capacity $K=5$ (census $N \approx 10,000$), average lifetime $L=4$, genome size $1\times10^{8}$, recombination rate $1\times10^{-9}$, and drawing $/sigma$ values from a uniform distribution bounded by 0.2 and 4. To speed up the simulations and limit memory overhead we used a mutation rate of 0 in SLiM and later applied mutations to the coalesced tree sequence with msprime's 'mutate( )' function. Because msprime assumes that timesteps are in units of generations, we divided the per-generation rate of $1\times10^{-8} mut/site/gen$ by the generation time estimated for each value of $\sigma$ (see 'Genealogical Parameters' below) to convert the rate to units of mutations/site/timestep. We then verified that this procedure produced the correct number of mutations by comparing a subset of simulations with SLiM-generated mutations (which are applied only at mating events) with mutations added by msprime (Supplementary Figure 1). Simulations were run until all extant individuals shared a common ancestor within the simulation (i.e. the tree sequence had coalesced).

@@

@@

\subsection{Demographic Modeling}
We fit single-population demographic models to the site frequency spectra of 20 randomly sampled individuals from each spatial SLiM simulation with the program Stairwayplot \citep{Liu2015}. Site frequency spectra used for input data were calculated in scikit-allel \citep{Miles2017}. 100 bootstrap replicates were generated for each simulation by resampling over sites. Models were then fit across all bootstrap replicates using default settings in Stairwayplot and the median estimate of $Ne$ per generation was used to represent the output of each simulation. The resulting inferred population size histories were plotted with line colors scaled to $\sigma$ to visualize how dispersal scaling impacts estimates of demographic history.

@@

\subsection{Summary Statistics}
As expected from theory based on the Wright-Malécot model \citep{Maruyama1972}, limited dispersal increases the coalescence time of a set of randomly-sampled individuals relative to a random-mating model. This is reflected in sequence data as a larger number of segregating sites and inflated estimates of expected heterozygosity ($\pi$, Figure 2). In practice the difference in segregating sites and expected heterozygosity in our simulations is negligible above neighborhood sizes of 65 ($\sigma$=1). Conversely observed heterozygosity (the fraction of heterozygous individuals) is lower than the random mating expectation when dispersal is low, and as a consequence the inbreeding coefficient ($F_{is}$) is positive. This is the continuous space version of the Wahlund effect \citep{Wahlund1928} - a deficit of heterozygous individuals when samples are drawn from subdivided populations. Similar observations of positive $F_{is}$ have been observed from stepping-stone simulations in \citep{Shirk2014} and \citep{Neel2013}.

@@

@@

\begin{figure*}[p]
\centering
\includegraphics[width=\textwidth]{sumstats_by_sigma.pdf}
\caption{Summary statistics from spatial and random-mating SLiM simulations. Statistics are calculated from the genotype matrix of 60 randomly sampled individuals in each simulation.}
\label{fig:spectrum}
\end{figure*}

@@

\afterpage{\clearpage}
\begin{figure*}[p]
\centering
\includegraphics[width=\textwidth]{gwas_summary_nsig_qqplots_logX.pdf}
\caption{Impacts of spatially correlated phenotypes and isolation by distance on linear regression GWAS. Colors represent different phenotype distributions as indicated in the figure legend. A: example phenotype distributions and sampling density. B: proportion of total variance explained by the first 10 PC axes, by neighborhood size. C: numbers of significant SNPs after FDR correction for linear-model GWAS conducted with (top) or without (bottom) PC covariates. D: Quantile-quantile plots of observed versus $p$ values relative to those expected from a uniform distribution for neighborhood sizes under 65 ($\sigma$=1). The dotted lines show the 95\% confidence region assuming binomial sampling and an alpha level of 0.05. Points above the 1:1 line reflect inflated -log10(p).}
\label{fig:spectrum}
\end{figure*}

>>>>

\begin{abstract}
Individuals exist in continuous space, but standard models in population genetics are based on discrete randomly-mating populations exchanging migrants. As the availability of population-level whole genome data allows inference of increasingly fine-scale patterns of ancestry in many species, models incorporating realistic demographic and spatial processes are needed to accurately describe spatial structure and control for its impacts on analyses of selection and demography. Here we implement a forward-time simulation of evolution in continuous space and use it to study the impacts of dispersal and density-dependent competition on population genetic summary statistics, demographic inference, and association studies. Low dispersal slows the geographic spread of ancestry and increases branch lengths corresponding to mid-frequency alleles because of slow coalescence among distant groups of individuals. As a result demographic inference from the site frequency spectrum is biased towards inferring a turbulent past demographic history and a recent decline in populations when neighborhood sizes are below 20. For many common population genetic summary statistics the variation associated with different sampling strategies is larger than that observed across the range of dispersal distances we simulated, pointing to the need to carefully account for the spatial distribution of sampling when interpreting results. Last we show that spatially correlated environments cause genome-wide signals of association with purely environmental phenotypes in linear-regression GWAS, and this bias is not fully corrected by regressing out principal components positions during the analysis.  
\end{abstract}

@@

In most sexual organisms individuals mate with other nearby individuals, leading to a positive correlation between genetic and geographic distances. This pattern of "isolation by distance" \citep{Wright1943} is one of the most widely replicated empirical findings in population genetics \citep{Chen2017,Jay2012,Sharbel2000}, but is only heuristically approximated by existing forward-time models of molecular evolution based on discrete demes exchanging migrants. As the availability of population-level whole-genome data allows inference of increasingly fine-scale patterns of ancestry in many species, models incorporating biologically realistic demographic and spatial processes are needed to accurately describe continuous spatial structure and control for its impacts on analyses of selection and demography. This situation is particularly clear in human population genetics, where individual ancestries are often interpreted as mixtures of discrete populations in part because of modeling assumptions underlying discrete clustering algorithms and association analyses are often strongly dependent on statistical corrections for population stratification \citep{Berg2018,Bulik-Sullivan2015,Young2018,Heckerman2016,Kang2010,Sohail2018}.

The best-studied approaches to population genetics in continuous space were developed by \citep{Wright1943} and \citep{Malecot1948}, who derived expressions for genetic differentiation in continuous space assuming Poisson distributed numbers of offspring and independent dispersal among individuals. A key finding of Wright's model is that many important aspects of continuous populations can be described in terms of "neighborhood size" -- the number of potential mates for an individual in a given generation, defined as $4\pi\sigma^2d$, where $\sigma$ is the average dispersal distance and $d$ is population density. \cite{Maruyama1972} found that the rate of decline in genetic diversity in a 2-dimensional continuous population shifts between distinct regimes when $D\sigma^2$ is greater or less than 1, and proposed that this had the important implication that most population genetic expectations for randmoly mating populations could be applied to continuously distributed populations with relatively little error.

Though some aspects of continuous populations are fairly well described by the Wright-Malécot models, \cite{Felsenstein1975} showed that the assumptions of independent dispersal and Poisson distributed offspring that are the basis of these models are incompatible. Over time, a population meeting them will clump into a small number of geographic clusters occupying only a part of the available range. Although real populations are often clumped on landscapes due to factors like varying habitat quality and competition among species, the Wright and Malécot models produce much more extreme clumping than is observed in real organisms and fail to account for the density-dependent declines in population growth rate that are widely observed in real populations (CITE).

One method for modeling populations in continuous space is then to assume the existence of a grid of discrete randomly-mating populations connected by migration, which prevents clustering by forcing all regions to be occupied in every generation. Among many other important results drawn from this class of "lattice" or "stepping stone" models, \cite{Rousset1997} showed that the slope of the a linear regression of genetic differentiation ($F_{st}$) against the logarithm of spatial distance is an estimate of neighborhood size. Though good approximations of continuous structure given high dispersal (Barton(?)), these models are not truly continuous, force a uniform realized population density across landscapes, and limit investigation of spatial structure below the level of the deme. An alternative method is to model the geographic spread of ancestry backwards in time through a diffusion approximation -- an approach that has recently made significant progress in modeling both dispersal and demographic parameters \citep{Barton2010,Kelleher2014,Ringbauer2017,Ringbauer2018}. However these models currently lack a well-described forward-time analog that produces biologically realistic spatial patterns and are subject to the same difficulty in incorporating selection and other biological processes as all other reverse time methods (CITE??).

We took a direct approach to the clustering problem of classical forward-time models by incorporating density dependence into the model. By scaling the probability of survival in each timestep to local population density we shift reproductive output towards regions of low-density and prevent populations from clustering. A similar approach was taken previously by \citep{Doebeli2003} who used an individual based model with continuous space and density dependent fitness to study the probability of speciation along continuous environmental gradients. However to our knowledge all previous implementations of continuous space models have focused on a small number of genetic loci as the unit of analysis, which limits the ability to investigate the impacts of continuous space on genome-wide genetic variation as is now routinely sampled from real organisms.

Here we describe an implementation of an individual-based model in continuous space that incorporates overlapping generations, a Gaussian dispersal kernel, and density-dependent fitness. The model scales to chromosome-scale alignments across tens of thousands of individuals, and outputs the full genealogy and recombination history of all final-generation individuals. We use this simulation to describe how sampling strategy interacts with isolation by distance to cause systematic variation in population genetic summary statistics typically analyzed under discrete population models. We then examine how the fine-scale spatial structures occurring under limited dispersal impact demographic inference from the site frequency spectrum; finding that inflation of branch lengths for mid-frequency alleles associated with slow coalescence of distant individuals can bias analyses towards inferring turbulent demographic histories and recent declines in population size.

Last, we use our model to study the impacts of isolation by distance in continuous space on genome-wide association studies (GWAS). Variation in ancestry proportions between case and control cohorts is expected to cause inflation of test statistics at SNPs with different allele frequencies among populations \citep{Price2006}. Because most phenotypes are influenced by the environment and environmental factors are often spatially correlated, subtle spatial structure within continuous populations can also deflate $p$ values in GWAS of quantitative traits \citep{Mathieson2012}. These issues are particularly prominent for polygenic traits, because the deflated $p$ values expected from polygenic phenotype correlations are superficially similar to those expected from population structure \citep{Bulik-Sullivan2015,Young2018}. However most studies examining the effects of stratification on GWAS have focused on simulating phenotypes for real genotype data \citep{Price2006,Bulik-Sullivan2015} or generating simulated genotypes by sampling from static allele frequency distributions \citep{Bulik-Sullivan2015,Heckerman2016}. These approaches address application to human data but are not based on a true generative model and so provide limited insight into the processes generating stratification bias in association analyses. Here we simulate both the underlying genealogy and phenotypes of a continuously distributed population and seek to identify regions of parameter space -- i.e. the strength of isolation by distance and the spatial distribution of environmental effects on phenotypes -- in which common methods of stratification bias corrections in GWAS are subject to bias.

%[move some of this to methods, rewrite to avoid repitition w above]  A common method of correcting for population structure in GWAS is to run a principal components analysis on the genotype matrix and include PC coordinates as covariates in a linear regression for each SNP \citep{Price2006}. Mixed model GWAS employs a different statistical framework but takes a similar approach by incorporating a kinship matrix describing pairwise genetic relatedness among samples as a random effect in a linear mixed model \citep{Yu2006,Kang2008}. In both approaches the key question is whether or not the ancestry components of the model adequately describe "background" levels of genetic variation in the sample relative to any confounding effects of ancestry and phenotype. Recent studies in European and British populations \citep{Kerminen2018,Berg2018,Sohail2018} have found evidence of residual population stratification in effect size estimates from large-scale GWAS employing both linear and mixed model approaches, suggesting that existing structure corrections may be inadequate. For traits controlled by a few large effect loci this confounding effect is likely to be relatively small, but for analyses of complex traits with thousands of putatively associated SNPs the potential for confounding is high. Here we simulate a range of spatially correlated phenotypes for simulated individuals and seek to identify conditions under which existing GWAS structure corrections are subject to bias caused by continuous population structure.

\section{Materials and Methods}
\label{sec:materials:methods}

@@

To isolate spatial effects from other components of the model such as overlapping generations, increased variance in reproductive success, and density-dependent fitness, we  also implemented a second version of the SLiM simulation in which mates are selected at random and offspring disperse to a random location on the landscape. For all simulations the full genealogy and recombination history of final-generation individuals were stored as tree sequences \citep{Kelleher2018}.

@@

We ran 400 simulations for the spatial and random-mating SLiM models on a square landscape of size 50 with per-unit carrying capacity $K=5$ (census $N \approx 10,000$), average lifetime $L=4$, genome size $1\times10^{8}$, recombination rate $1\times10^{-9}$, and drawing $/sigma$ values from a uniform distribution bounded by 0.2 and 4. To speed up the simulations and limit memory overhead we used a mutation rate of 0 in SLiM and later applied mutations to the coalesced tree sequence with msprime's 'mutate( )' function. Because msprime assumes that timesteps are in units of generations, we divided the per-generation rate of $1\times10^{-8} mut/site/gen$ by the generation time estimated for each value of $\sigma$ (see 'Genealogical Parameters' below) to convert the rate to units of mutations/site/timestep. We then verified that this procedure produced the correct number of mutations by comparing a subset of simulations with SLiM-generated mutations (which are applied only at mating events) with mutations added by msprime (Supplementary Figure 1). Simulations were run for 1.6 million timesteps (approximately 30N generations), or until all extant individuals shared a common ancestor within the simulation (i.e. the tree sequence had coalesced).

@@

\subsection{Sampling}
Unlike reverse time approaches our model records the genealogy and sequence variation of the complete population, but in real populations genotypes are only observed from a subset of individuals. We modeled three sampling strategies meant to approximate common data collection methods in empirical genetic studies (Figure 1). "Random" sampling selects individuals at random from across the full landscape. "Point" sampling selects individuals proportional to their distance from four equally spaced points on the landscape. Last, "midpoint" sampling selects individuals in proportion to their distance from the middle of the landscape. In practice random sampling is rarely conducted because of the difficulty of accessing all parts of a species range and the inefficiency of collecting only single individuals at all sites. Point and midpoint sampling strategies are both common, particularly for species that require traps or collecting stations to be set up and dismantled at every site.

@@

\subsection{Demographic Modeling}
We fit single-population demographic models to the site frequency spectra of 20 individuals from each spatial SLiM simulation with the program Stairwayplot \citep{Liu2015}. This analysis was replicated across random, point, and midpoint sampling strategies. Site frequency spectra used for input data were calculated in scikit-allel \citep{Miles2017}. 100 bootstrap replicates were generated for each simulation by resampling over sites. Models were then fit across all bootstrap replicates using default settings in Stairwayplot and the median estimate of $Ne$ per generation was used to represent the output of each simulation. The resulting inferred population size histories were plotted with line colors scaled to $\sigma$ to visualize how dispersal scaling impacts estimates of demographic history.

@@

\subsection{Summary Statistics}
\subsubsection{Diversity}
As expected from theory based on the Wright-Malécot model \citep{Maruyama1972}, limited dispersal increases the coalescence time of a set of randomly-sampled individuals relative to a random-mating model. This is reflected in sequence data as a larger number of segregating sites and inflated estimates of expected heterozygosity ($\pi$, Figure 2). In practice the difference in segregating sites and expected heterozygosity in our simulations is negligible above neighborhood sizes of 65 ($\sigma$=1). Conversely observed heterozygosity (the fraction of heterozygous individuals) is lower than the random mating expectation when dispersal is low, and as a consequence the inbreeding coefficient ($F_{is}$) is positive. This is the continuous space version of the Wahlund effect \citep{Wahlund1928} - a deficit of heterozygous individuals when samples are drawn from subdivided populations. Similar observations of positive $F_{is}$ have been observed from stepping-stone simulations in \citep{Shirk2014} and \citep{Neel2013}.

@@

\subsubection{Site Frequency Spectrum}

\subsubsection{Haplotype Block Lengths}

\subsubsection{Spatial Correlations}

@@

\begin{figure*}[p]
\centering
\includegraphics[width=\textwidth]{sumstats_by_neighbors.pdf}
\caption{[needs some polishing for labels etc] Summary statistics from spatial and random-mating SLiM simulations. Statistics are calculated from the genotype matrix of 60 randomly sampled individuals in each simulation.}
\label{fig:spectrum}
\end{figure*}

@@

\afterpage{\clearpage}
\begin{figure*}[p]
\centering
\includegraphics[width=\textwidth]{gwas_summary_nsig_qqplots_loglog.pdf}
\caption{Impacts of spatially correlated phenotypes and isolation by distance on linear regression GWAS. Colors represent different phenotype distributions as indicated in the figure legend. A: example phenotype distributions and sampling maps, with colors scaled to phenotype values. B: proportion of total variance explained by the first 10 PC axes, by neighborhood size. C: numbers of significant SNPs after FDR correction for linear-model GWAS conducted with (top) or without (bottom) PC covariates. D: Quantile-quantile plots of observed versus $p$ values relative to those expected from a uniform distribution. The dotted lines show the 95\% confidence region assuming binomial sampling and an alpha level of 0.05. Points above the 1:1 line reflect inflated -log10(p).}
\label{fig:spectrum}
\end{figure*}
